Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-10-11 SUBMISSION_03/article/article.tex
Commit 2024-10-12 01:08

The cloud-free reference values discussed above are based on the
climatological AOD and WC; hence they cannot describe accurately the
long-term variation of \(\text{GHI}_\text{ref}\) due to long-term
changes in the two atmospheric constituents, mainly AOD. As reported by
\citet{Natsis2023}, there is a long-term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for cloud-free
data was attributed to long-term changes in aerosol effects. Therefore,
an adjustment of the \(\text{GHI}_\text{ref}\) during the period of
study was made using RTM simulations to account for the long-term
variations of the AOD. High quality AOD data with a Cimel sun-photometer
at Thessaloniki start only in 2003, while spectral AOD measurements
using direct-sun observations with a MKIII Brewer spectrophotometer are
available for the period 1997 -- 2017. The Brewer AOD data are taken
sporadically during each day and are less dense compared to the AERONET
data. By comparing monthly AOD data at \(340\,\text{nm}\) of the two
instruments for the common periods of operation we adjusted the Brewer
data and filled the missing months of the Cimel record with Brewer data.
Using the extended monthly time series of AOD at \(340\,\text{nm}\), as
well as monthly climatological values of the Ångström exponent and
constant WC of \(15.6\,\text{mm}\) derived from the Cimel record, we
simulated with the RTM the cloud-free GHI at SZA of \(55^\circ\) for
each month in the period 1997 -- 2023. The SZA of \(55^\circ\) was
chosen as representative of all days in the year to get a rough estimate
of the annually averaged change in cloud-free GHI. A second-degree
polynomial fit was applied to the simulated yearly averaged GHI to
derive the long-term change in GHI due to aerosols: \begin{equation}
\Delta(\text{GHI}) [\%] = -12170 + 12.05 \cdot x + -0.002981 \cdot x^2 \label{eq:AODchange}
\end{equation} where \(y\) is the date as a decimal fraction of the
year. The resulting change in \(\text{GHI}_\text{ref}\) ranges between
\(-1.88\) in 1994 and \(-0.19\,\%\) in the end of 2023, peaking at
\(0.23\,\%\) in the mid-2020. Finally, we applied these long-term
changes of Equation\nobreakspace{}\ref{eq:AODchange} to the
\(\text{GHI}_\text{ref}\), to create a more realistic representation of
the cloud-free irradiance for the entire period of study. For the period
1994 -- 1996 where no AOD data are available, we assumed that the
changes in GHI follow the same polynomial fit.

@@

One of the goals of this study was to quantify the OI related to CEs. A
key issue for achieving this goal is to define a threshold for the CE
identification, representative of the cloud-free irradiance at the time
of each GHI measurement. This depends on the selection of the
appropriate atmospheric parameterization for the RTM simulations. The
implementation of the long-term changes in GHI due to AOD, discussed in
section\nobreakspace{}\ref{rtmcs}, allows capturing a large part of the
natural variability of cloud-free GHI. However, the short-term
variability of AOD cannot be taken adequately into account when using
monthly values in the model simulations. We tried different approaches
to strengthen the robustness of the methodology and to compensate for
the limited accuracy of the RTM input data and the unpredictable natural
variability of the atmosphere.

First, we evaluated the performance of the modeled
\(\text{GHI}_\text{ref}\) in relation to the measured GHI for different
levels of atmospheric clearness, by using in the RTM the monthly
climatological AOD and WC, less their respective standard deviations.
These values represent typical atmospheres in Thessaloniki with
lower-than-average load of aerosols and humidity, which are the main
factors that attenuate the GHI, in the absence of clouds. With this
approach, the simulated \(\text{GHI}_\text{ref}\) should generally be
greater than the measured GHI when aerosols are more abundant. The
correlation of the \(\text{GHI}_\text{ref}\) with the GHI was tested for
a subset of 513 cloud-free days with more than \(80\,\%\) data
availability (Figure\nobreakspace{}\ref{fig:validation-GHI}). The
scatter plot shows a good agreement between the modeled and the measured
GHI with a spread that arises mainly from the short-term variability of
AOD and WC. The linear regression reveals a positive bias of
\(18.4\,\text{W}/\text{m}^2\) and a slope very close to unity. To
achieve a robust and clear distinction of the CE cases, we applied to
the \(\text{GHI}_\text{ref}\) an additional offset of
\(25\,\text{W}/\text{m}^2\).

@@

\caption{Relation of excess irradiation of CE groups with their duration for Thessaloniki in the period 1994 -- 2023. The logarithmic color scale denotes the frequency of the respective groups. Note that 19 pints with duration higher than 60 min are not shown.}\label{fig:group-2d}
\end{figure}

>>>>

The cloud-free reference values discussed above are based on the
climatological AOD and WC; hence they cannot describe accurately the
long-term variation of \(\text{GHI}_\text{ref}\) due to long-term
changes in the two atmospheric constituents, mainly AOD. As reported by
\citet{Natsis2023}, there is a long-term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for cloud-free
data was attributed to long-term changes in aerosol effects. Therefore,
an adjustment of the \(\text{GHI}_\text{ref}\) during the period of
study was made using RTM simulations to account for the long-term
variations of the AOD. High quality AOD data with a Cimel sun-photometer
at Thessaloniki start only in 2003, while spectral AOD measurements
using direct-sun observations with a MKIII Brewer spectrophotometer are
available for the period 1997 -- 2017. The Brewer AOD data are taken
sporadically during each day and are less dense compared to the AERONET
data. By comparing monthly AOD data at \(340\,\text{nm}\) of the two
instruments for the common periods of operation we adjusted the Brewer
data and filled the missing months of the Cimel record with Brewer data.
Using the extended monthly time series of AOD at \(340\,\text{nm}\), as
well as monthly climatological values of the Ångström exponent and
constant WC of \(15.6\,\text{mm}\) derived from the Cimel record, we
simulated with the RTM the cloud-free GHI at SZA of \(55^\circ\) for
each month in the period 1997 -- 2023. The SZA of \(55^\circ\) was
chosen as representative of all days in the year to get a rough estimate
of the annually averaged change in cloud-free GHI. A second-degree
polynomial fit was applied to the simulated yearly averaged GHI to
derive the long-term change in GHI due to aerosols: \begin{equation}
\Delta(\text{GHI}) [\%] = -12153.97 + 12.0304 \cdot y -0.002977 \cdot y^2 \label{eq:AODchange}
\end{equation} where \(y\) is the year in decimal form. The resulting 
change in \(\text{GHI}_\text{ref}\) ranges between \(-2\,\%\) in 1994
and \(0.05\,\%\) in the end of 2023, peaking at \(0.09\,\%\) in the
mid-2020. Finally, we applied these long-term changes of
Equation\nobreakspace{}\ref{eq:AODchange} to the
\(\text{GHI}_\text{ref}\), to create a more realistic representation of
the cloud-free irradiance for the entire period of study. For the period
1994 -- 1996 where no AOD data are available, we assumed that the
changes in GHI follow the same polynomial fit.

@@

One of the goals of this study was to quantify the OI related to CEs. A
key issue for achieving this goal is to define a threshold for the CE
identification, representative of the cloud-free irradiance at the time
of each GHI measurement. This depends on the selection of the
appropriate values of the input parameters for the RTM simulations. The
implementation of the long-term changes in GHI due to AOD, discussed in
section\nobreakspace{}\ref{rtmcs}, allows capturing a large part of the
natural variability of cloud-free GHI. However, the short-term
variability of AOD cannot be taken adequately into account when using
monthly values in the model simulations. We tried different approaches
to strengthen the robustness of the methodology and to compensate for
the limited accuracy of the RTM input data and the unpredictable natural
variability of the atmosphere.

First, we evaluated the performance of the modeled
\(\text{GHI}_\text{ref}\) in relation to the measured GHI for different
levels of atmospheric clearness, by using in the RTM the monthly
climatological AOD and WC, less their respective standard deviations.
These values represent typical atmospheres in Thessaloniki with
lower-than-average load of aerosols and humidity, which are the main
factors that attenuate the GHI, in the absence of clouds. With this
approach, the simulated \(\text{GHI}_\text{ref}\) should generally be
greater than the measured GHI when aerosols are more abundant. The
correlation of the \(\text{GHI}_\text{ref}\) with the GHI was tested for
a subset of 513 cloud-free days with more than \(80\,\%\) data
availability (Figure\nobreakspace{}\ref{fig:validation-GHI}). The
scatter plot shows a good agreement between the modeled and the measured
GHI with a spread that arises mainly from the short-term variability of
AOD and WC. The linear regression reveals a positive bias of
\(18.4\,\text{W}/\text{m}^2\) and a slope very close to unity. To
achieve a robust and clear distinction of the CE cases, we applied to
the \(\text{GHI}_\text{ref}\) an additional offset of
\(25\,\text{W}/\text{m}^2\). The long-term difference between the
measured GHI and the \(\text{GHI}_{\text{ref}}\) for the same cloud-free
days has a trend of \(0.2\,\text{W}/\text{m}^2\) per year which further
supports the validity of the produced long-term
\(\text{GHI}_\text{ref}\).

@@

\caption{Relation of excess irradiation of CE groups with their duration for Thessaloniki in the period 1994 -- 2023. The logarithmic color scale denotes the frequency of the respective groups. Note that 19 points for groups with duration higher than 60 min are not shown.}\label{fig:group-2d}
\end{figure}
